---
title: "Breast cancer microbiome case/control signal implantation benchmark analysis using SIMBA"
author:
  - name: "Mani Arumugam"
    affiliation: "University of Copenhagen"
    email: "arumugam@sund.ku.dk"
date: "Date last modified: 16-07-2026"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Breast cancer microbiome case/control signal implantation benchmark analysis using SIMBA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

This document is a step-by-step walkthrough for reproducing the breast cancer status detectability analysis in the manuscript. The analysis starts from a single `phyloseq` object and proceeds through four main stages:

1. prepare the count table and sample metadata;
2. generate signal-implantation simulations with a patched version of SIMBA;
3. benchmark multiple differential abundance methods on the simulated data; and
4. compare the real-data results with simulation-derived detectability boundaries.

The goal is not only to reproduce the figures in the manuscript, but also to make the logic of the workflow transparent to anyone who wants to rerun or adapt it to a comparable dataset.

# Before you start

## Required inputs

To reproduce the analysis, you will need:

- a `phyloseq` object containing the taxon count table and sample metadata;
- installation of the patched SIMBA fork from https://github.com/microbiomix/SIMBA
  - (use `devtools::install_github(repo="https://github.com/microbiomix/SIMBA")`)
- the companion helper scripts `helper_functions.R`, `helper_detectability.R`; and
- a writable temporary directory for the HDF5 simulation file and output figures.

## What this document assumes

This workflow assumes that the starting `phyloseq` object contains:

- a count-based OTU table;
- a sample-level identifier that can be used as row names for metadata; and
- the clinical variables needed for the benchmarked models.

In the present breast cancer analysis, the main outcome is `DiseaseStatus.binary`, and the covariate-adjusted benchmark adjusts for `Age`.

# User configuration

Edit this chunk before running the document. The paths below are examples and should be replaced with the locations on your own system.

```{r config}
# Paths that users should modify -------------------------------------------

# Making sure SIMBA is available

# PREFERRED OPTION:
# Installed via GitHub
# devtools::install_github(repo="https://github.com/microbiomix/SIMBA")
suppressWarnings(library(simbaR))

# Alternatively, use a local path to the patched SIMBA fork
# Use this if you have not installed simbaR from github
# Pass this to 'run_method_eval_parallel()' via simba.loc argument
simba.loc <- "/path/to/microbiomix/SIMBA"

# Directory containing helper_functions.R and sensitivity_functions.R
code.loc <- "/path/to/analysis/helpers"

# Writable directory for figure outputs
out.loc <- "/path/to/output/"

# Writable directory for simulation HDF5 files
simh5.loc <- "/path/to/output/"

dir.create(out.loc, recursive = TRUE, showWarnings = FALSE)

# Input phyloseq object
ps_rds <- "/path/to/ps.motus_raw.3.0.3.rds"

# Number of CPUs to use when parallelizing DA test benchmarks
n_cpus <- 8
```

```{r hidden-config, message=FALSE, warning=FALSE, include=FALSE}
simba.loc <- "."
code.loc <- "./"
out.loc <- "./figures/"
simh5.loc <- "/scratch/mjq180/SIMBA/"
ps_rds <- "/projects/arumugam/people/xhf865/Camila/BreastCancer/phyloseq/ps.motus_raw.3.0.3.rds"
n_cpus <- 121
```

# Load packages and helper functions

The benchmark relies on the patched SIMBA code together with the helper functions used to extract calibration summaries and generate the final figures.

```{r mysetup, message=FALSE, warning=FALSE}
suppressWarnings(library(tidyverse))
suppressWarnings(library(phyloseq))
suppressWarnings(library(dplyr))
suppressWarnings(library(purrr))
suppressWarnings(library(stringr))
suppressWarnings(library(tibble))
suppressWarnings(library(rhdf5))
suppressWarnings(library(ggplot2))
suppressWarnings(library(patchwork))
suppressWarnings(library(ANCOMBC))
suppressWarnings(library(metagenomeSeq))
suppressWarnings(library(LinDA))
suppressWarnings(library(DESeq2))
suppressWarnings(library(edgeR))
suppressWarnings(library(compositions))
suppressWarnings(library(DT))

# Load companion helper functions used throughout the workflow
source(file.path(code.loc, "helper_functions.R"))
source(file.path(code.loc, "helper_detectability.R"))
```

# Load and prepare the phyloseq object

## Global analysis settings

These settings define the minimum prevalence filter used throughout the analysis and the significance threshold used in the main benchmark summaries. They also define the number of reps made in the simulation, and how many of them are used for benchmarking.

```{r globals}
REPS_MADE <- 50
REPS_USED <- 50
MIN_BENCHMARK_PREVALENCE <- 0.10
MIN_DISCOVERY_PREVALENCE <- 0.10
ALPHA <- 0.10
param_set <- paste("alpha", ALPHA, "prev", MIN_BENCHMARK_PREVALENCE, "reps", REPS_USED, sep = "_")
```

## Read the phyloseq object and apply prevalence filtering

We begin from the downloaded `phyloseq` object and remove taxa with prevalence below the chosen threshold.

```{r load-phyloseq}
ps_orig <- readRDS(ps_rds)
ps <- filter_taxa_by_prevalence(ps_orig, MIN_BENCHMARK_PREVALENCE)
```

## Prepare metadata for SIMBA

SIMBA expects a plain metadata data frame with samples in rows. In this analysis we keep only the variables needed for benchmarking and covariate-adjusted method comparisons.

```{r prepare-metadata}

meta <- sample_data(ps) %>%
  unclass() %>%
  as.data.frame() %>%
  select(
    participant_id,
    DNA_ID,
    DiseaseStatus.binary,
    Age,
    bmi_calculated
  )

meta$Individual_ID <- as.character(meta$participant_id)
meta$participant_id <- NULL
meta$Timepoint <- 1
rownames(meta) <- meta$DNA_ID
meta$DNA_ID <- NULL

sample_data(ps) <- meta
```

## Prepare the count table for SIMBA

The signal implantation framework uses integer counts. We therefore extract the OTU table from `phyloseq`, keep taxa as rows, and round to integer counts.

```{r prepare-counts}
feat <- otu_table(ps) %>%
  unclass() %>%
  as.data.frame()

rownames(feat) <- taxa_names(ps)
feat[] <- lapply(feat, function(x) as.integer(round(x)))
```


# Define simulation parameters

We now specify the filtering parameters used internally by SIMBA, the proportion of implanted marker taxa, class balance, subset size, the abundance-shift grid and the number of simulation repeats.

In this benchmark, `ab.scale` is the primary implanted effect-size parameter. We set `prev.scale = 0`, meaning that signals are implanted as abundance differences without additional prevalence shifts.

```{r parameters}
# SIMBA filtering parameters stored in the HDF5 file
ab.cutoff <- 1e-04
prev.cutoff <- MIN_BENCHMARK_PREVALENCE
log.n0 <- 1e-05

# Simulation design
prop.markers <- 0.1
class.balance <- 101 / 261
subset_size <- 261
sim_reps <- REPS_MADE

# Implanted effect sizes
ab.scale <- seq(1, 10, 0.5)

prev.scale <- 0
```

# Generate the signal-implantation simulations

The patched SIMBA workflow preserves the original sample-to-metadata mapping and allows covariates to remain attached to the same samples during benchmarking. This is crucial when the downstream methods use covariate adjustment.

## Create the simulation HDF5 file

Because labels are fixed in this benchmark, we only need one testing-index repetition for each subset size.

```{r create-simulation, message=FALSE}

force_simulation <- FALSE
seed <- 1234
param_set1 <- paste("prev", MIN_BENCHMARK_PREVALENCE, "reps", REPS_MADE, sep = "_")
sim.implant <- file.path(simh5.loc, paste0("BC_", param_set1, "_seed_", seed, "_implantation.h5"))
if (!file.exists(sim.implant) || force_simulation) {
  set.seed(seed)
  if (file.exists(sim.implant)) {
    file.remove(sim.implant)
  }
  system.time(
    create.data.simulation(
      feat = feat,
      meta = meta,
      sim.location = sim.implant,
      sim.type = "cross-section",
      sim.method = "resampling",
      filt.params = list(
        ab.cutoff = as.numeric(ab.cutoff),
        prev.cutoff = prev.cutoff,
        log.n0 = as.numeric(log.n0)
      ),
      sim.params = list(
        ab.scale = ab.scale,
        prev.scale = prev.scale,
        prop.markers = prop.markers,
        conf = NULL,
        class.balance = class.balance,
        feature.type = "all",
        fixed.labels = TRUE,
        factorize.metadata = FALSE,
        label.col = "DiseaseStatus.binary",
        case.value = 2,
        control.value = 1,
        repeats = sim_reps
      )
    )
  )
  system.time(
    create.test.idx(
      sim.location = sim.implant,
      subsets = subset_size,
      repetitions = 1,
      fixed.labels = TRUE
    )
  )

}
```


Compare sample-anchored labels that are stored

```{r compare-labels-in-test-idx,, eval=FALSE, include=FALSE}
library(rhdf5)

g <- "ab1_prev1_rep1"
subset_name <- "subset_261"

grp <- h5read(sim.implant, g)
params <- h5read(sim.implant, "simulation_parameters")

# true labels written by simulate.resampling()
true_label <- grp$labels
sample_names <- grp$sample_names
names(true_label) <- sample_names

# metadata with preserved sample IDs
meta2 <- h5read(sim.implant, "original_data/metadata")
meta2 <- as.data.frame(meta2, stringsAsFactors = FALSE)
rownames(meta2) <- meta2[["__sample_id"]]
#meta[["__sample_id"]] <- NULL

# stored test_idx object
info.mat <- grp$test_idx[[subset_name]]
rownames(info.mat) <- params[["test_idx"]][["names"]]

# map index positions to sample names
stored_sample_names <- sample_names[info.mat["rep1", ]]

# compare stored label row to true labels / metadata labels for rep1
data.frame(
  sample = stored_sample_names,
  stored_label = info.mat["label", ],
  true_label = true_label[stored_sample_names],
  meta_label = meta2[stored_sample_names, "DiseaseStatus.binary", drop = TRUE]
)
```

A quick look at the simulation groups confirms that the HDF5 file contains one group per implanted effect-size setting.

```{r list-simulation-groups}
temp <- rhdf5::h5ls(sim.implant, recursive = FALSE)
length(setdiff(temp$name, c("original_data", "simulation_parameters")))
setdiff(temp$name, c("original_data", "simulation_parameters"))[1:10]
rm(temp)
```

# Reality check for the simulations

Before running the benchmark, it is useful to verify that the implanted samples still resemble the original data structure. Principal coordinates analysis (PCoA) can be used as a simple visual check.

```{r pcoa-plots, collapse=TRUE, results='hold', fig.width=8, fig.height=7}
pcoa.plot(sim.implant, "ab2_prev1_rep1", "bray") +
  labs(title = "Signal implantation (ab.scale=1.5)")
pcoa.plot(sim.implant, "ab5_prev1_rep1", "bray") +
  labs(title = "Signal implantation (ab.scale=3)")
pcoa.plot(sim.implant, "ab11_prev1_rep1", "bray") +
  labs(title = "Signal implantation (ab.scale=5)")

```

As `ab.scale` increases, simulated samples are moving farther away from real samples. This is as expected and confirms that our label preservation patch works.

# Verify our patched SIMBA

Let's make sure that our modifications do not behave strange using some more sanity checks.

Let us also check that the feature labels and sample labels are preserved.

```{r compare-original-with-sim, collapse=TRUE, results='hold'}

# ------------------------------------------------------------
# Sanity check: labels and abundances are preserved across
#   1) phyloseq object `ps`
#   2) sim.implant/original_data
#   3) sim.implant implanted group ab1_prev1_rep1
# ------------------------------------------------------------

g <- "ab1_prev1_rep1"

# ---- 1. Extract matrix from phyloseq and ensure taxa are rows ----
ps_mat <- as(otu_table(ps), "matrix")
if (!taxa_are_rows(ps)) {
  ps_mat <- t(ps_mat)
}

# ---- 2. Read original_data from HDF5 ----
orig_feat <- h5read(sim.implant, "original_data/filt_features")
orig_feat_names <- h5read(sim.implant, "original_data/filt_feature_names")
orig_sample_names <- h5read(sim.implant, "original_data/sample_names")

rownames(orig_feat) <- orig_feat_names
colnames(orig_feat) <- orig_sample_names

# ---- 3. Read one implanted group from HDF5 ----
grp <- h5read(sim.implant, g)
sim_feat <- grp$features
rownames(sim_feat) <- grp$feature_names
colnames(sim_feat) <- grp$sample_names

# ---- 4. Check labels before reordering ----
cat("Feature labels before reordering:\n")
cat("  ps vs original_data: ", isTRUE(all.equal(rownames(ps_mat), rownames(orig_feat))), "\n")
cat("  ps vs implanted    : ", isTRUE(all.equal(rownames(ps_mat), rownames(sim_feat))), "\n")
cat("  original vs implant: ", isTRUE(all.equal(rownames(orig_feat), rownames(sim_feat))), "\n\n")

cat("Sample labels before reordering:\n")
cat("  ps vs original_data: ", isTRUE(all.equal(colnames(ps_mat), colnames(orig_feat))), "\n")
cat("  ps vs implanted    : ", isTRUE(all.equal(colnames(ps_mat), colnames(sim_feat))), "\n")
cat("  original vs implant: ", isTRUE(all.equal(colnames(orig_feat), colnames(sim_feat))), "\n\n")

# ---- 5. Reorder all matrices by common feature and sample names ----
common_features <- sort(Reduce(intersect, list(
  rownames(ps_mat), rownames(orig_feat), rownames(sim_feat)
)))
common_samples <- sort(Reduce(intersect, list(
  colnames(ps_mat), colnames(orig_feat), colnames(sim_feat)
)))

ps_sub   <- ps_mat[common_features, common_samples, drop = FALSE]
orig_sub <- orig_feat[common_features, common_samples, drop = FALSE]
sim_sub  <- sim_feat[common_features, common_samples, drop = FALSE]

cat("Dimensions after alignment:\n")
cat("  ps        :", paste(dim(ps_sub), collapse = " x "), "\n")
cat("  original  :", paste(dim(orig_sub), collapse = " x "), "\n")
cat("  implanted :", paste(dim(sim_sub), collapse = " x "), "\n\n")

# ---- 6. Confirm that names now match exactly after reordering ----
cat("Feature labels after reordering:\n")
cat("  ps vs original_data: ", isTRUE(all.equal(rownames(ps_sub), rownames(orig_sub))), "\n")
cat("  ps vs implanted    : ", isTRUE(all.equal(rownames(ps_sub), rownames(sim_sub))), "\n")
cat("  original vs implant: ", isTRUE(all.equal(rownames(orig_sub), rownames(sim_sub))), "\n\n")

cat("Sample labels after reordering:\n")
cat("  ps vs original_data: ", isTRUE(all.equal(colnames(ps_sub), colnames(orig_sub))), "\n")
cat("  ps vs implanted    : ", isTRUE(all.equal(colnames(ps_sub), colnames(sim_sub))), "\n")
cat("  original vs implant: ", isTRUE(all.equal(colnames(orig_sub), colnames(sim_sub))), "\n\n")

# ---- 7. Compare matrix values ----
cat("Matrix equality after alignment:\n")
cat("  ps vs original_data: ", isTRUE(all.equal(ps_sub, orig_sub)), "\n")
cat("  ps vs implanted    : ", isTRUE(all.equal(ps_sub, sim_sub)), "\n")
cat("  original vs implant: ", isTRUE(all.equal(orig_sub, sim_sub)), "\n\n")

cat("Maximum absolute difference:\n")
cat("  ps vs original_data: ", max(abs(ps_sub - orig_sub), na.rm = TRUE), "\n")
cat("  ps vs implanted    : ", max(abs(ps_sub - sim_sub), na.rm = TRUE), "\n")
cat("  original vs implant: ", max(abs(orig_sub - sim_sub), na.rm = TRUE), "\n\n")

# ---- 8. Optional: inspect first mismatch if values differ ----
if (!isTRUE(all.equal(ps_sub, sim_sub))) {
  idx <- which(ps_sub != sim_sub, arr.ind = TRUE)
  if (nrow(idx) > 0) {
    i <- idx[1, "row"]
    j <- idx[1, "col"]
    cat("First ps vs implanted mismatch:\n")
    cat("  feature :", rownames(ps_sub)[i], "\n")
    cat("  sample  :", colnames(ps_sub)[j], "\n")
    cat("  ps      :", ps_sub[i, j], "\n")
    cat("  implant :", sim_sub[i, j], "\n")
  }
}

```

Verify that the implanted markers are the only ones that changed.

```{r verify-implantation-integrity, collapse=TRUE, results='hold'}

g <- "ab15_prev1_rep1"

# ------------------------------------------------------------
# 1. Read original and implanted feature tables
# ------------------------------------------------------------

# original data
orig_feat <- h5read(sim.implant, "original_data/filt_features")
orig_feat_names <- h5read(sim.implant, "original_data/filt_feature_names")
orig_sample_names <- h5read(sim.implant, "original_data/sample_names")

rownames(orig_feat) <- orig_feat_names
colnames(orig_feat) <- orig_sample_names

# implanted group
grp <- h5read(sim.implant, g)
sim_feat <- grp$features
rownames(sim_feat) <- grp$feature_names
colnames(sim_feat) <- grp$sample_names

marker_idx <- grp$marker_idx
label <- grp$labels
names(label) <- grp$sample_names

# ------------------------------------------------------------
# 2. Align by feature and sample names
# ------------------------------------------------------------

common_features <- sort(intersect(rownames(orig_feat), rownames(sim_feat)))
common_samples  <- sort(intersect(colnames(orig_feat), colnames(sim_feat)))

orig_sub <- orig_feat[common_features, common_samples, drop = FALSE]
sim_sub  <- sim_feat[common_features, common_samples, drop = FALSE]

label_sub <- label[common_samples]

cat("Aligned dimensions:\n")
cat("  original :", paste(dim(orig_sub), collapse = " x "), "\n")
cat("  implanted:", paste(dim(sim_sub), collapse = " x "), "\n\n")

# ------------------------------------------------------------
# 3. Which features actually changed?
# ------------------------------------------------------------

# any count change at all in at least one sample
changed_feature <- rowSums(orig_sub != sim_sub, na.rm = TRUE) > 0
changed_features <- rownames(orig_sub)[changed_feature]

cat("Features listed in marker_idx      :", length(marker_idx), "\n")
cat("Features with any observed change  :", length(changed_features), "\n")
cat("Intersection(marker_idx, changed)  :", length(intersect(marker_idx, changed_features)), "\n")
cat("Markers with no observed change    :", length(setdiff(marker_idx, changed_features)), "\n")
cat("Changed but not in marker_idx      :", length(setdiff(changed_features, marker_idx)), "\n\n")

# show examples if something looks wrong
cat("First marker_idx entries:\n")
print(head(sort(marker_idx), 10))

cat("\nFirst changed features:\n")
print(head(sort(changed_features), 10))

cat("\nMarkers with no observed change:\n")
print(head(setdiff(marker_idx, changed_features), 10))

cat("\nChanged but not in marker_idx:\n")
print(head(setdiff(changed_features, marker_idx), 10))

# ------------------------------------------------------------
# 4. Summarize per-feature change magnitude
# ------------------------------------------------------------

qc_df <- tibble(
  feature = rownames(orig_sub),
  is_marker = rownames(orig_sub) %in% marker_idx,
  n_changed_samples = rowSums(orig_sub != sim_sub, na.rm = TRUE),
  mean_abs_diff = rowMeans(abs(sim_sub - orig_sub), na.rm = TRUE),
  sum_abs_diff = rowSums(abs(sim_sub - orig_sub), na.rm = TRUE),
  prevalence_orig = rowMeans(orig_sub > 0, na.rm = TRUE),
  prevalence_sim  = rowMeans(sim_sub > 0, na.rm = TRUE)
) %>%
  arrange(desc(sum_abs_diff), desc(n_changed_samples))

cat("\nTop features by total absolute change:\n")
print(qc_df %>% select(feature, is_marker, n_changed_samples, sum_abs_diff, mean_abs_diff) %>% head(20))

# ------------------------------------------------------------
# 5. Check expected class separation among markers
# ------------------------------------------------------------

# compare mean relative abundance by class before and after implantation
orig_rel <- sweep(orig_sub, 2, colSums(orig_sub), FUN = "/")
sim_rel  <- sweep(sim_sub, 2, colSums(sim_sub), FUN = "/")

pos_samples <- names(label_sub)[label_sub == 1]
neg_samples <- names(label_sub)[label_sub == -1]

marker_effect_df <- tibble(
  feature = marker_idx[marker_idx %in% rownames(sim_rel)],
  orig_mean_pos = rowMeans(orig_rel[marker_idx[marker_idx %in% rownames(sim_rel)], pos_samples, drop = FALSE], na.rm = TRUE),
  orig_mean_neg = rowMeans(orig_rel[marker_idx[marker_idx %in% rownames(sim_rel)], neg_samples, drop = FALSE], na.rm = TRUE),
  sim_mean_pos  = rowMeans(sim_rel [marker_idx[marker_idx %in% rownames(sim_rel)], pos_samples, drop = FALSE], na.rm = TRUE),
  sim_mean_neg  = rowMeans(sim_rel [marker_idx[marker_idx %in% rownames(sim_rel)], neg_samples, drop = FALSE], na.rm = TRUE)
) %>%
  mutate(
    orig_diff = orig_mean_pos - orig_mean_neg,
    # true implanted separation after simulation
    sim_diff  = sim_mean_pos - sim_mean_neg,
    abs_sim_diff = abs(sim_diff)
  ) %>%
  arrange(desc(abs_sim_diff))

cat("\nTop implanted markers by absolute class separation after implantation:\n")
print(marker_effect_df %>% select(feature, sim_mean_pos, sim_mean_neg, sim_diff, abs_sim_diff) %>% head(20))

# ------------------------------------------------------------
# 6. Quick pass/fail summary
# ------------------------------------------------------------

all_markers_changed <- setequal(intersect(marker_idx, rownames(orig_sub)), intersect(marker_idx, changed_features))
no_nonmarkers_changed <- length(setdiff(changed_features, marker_idx)) == 0

cat("\nQC flags:\n")
cat("  all markers changed at least somewhere : ", all_markers_changed, "\n")
cat("  no non-marker changed                  : ", no_nonmarkers_changed, "\n")
```

Verify that running the `apply.test()` function from SIMBA still works as expected. Nothing should be detected because we are testing `ab.scale=1` here.

```{r verify-eval-integrity, collapse=TRUE, results='hold'}

alpha <- 0.05
rep_col <- "rep_1"

g <- "ab1_prev1_rep1"
subset_name <- "subset_261"

grp <- rhdf5::h5read(sim.implant, g)
marker_idx_h5 <- grp$marker_idx

res <- apply.test(
  sim.location = sim.implant,
  group = g,
  subset = 261,
  norm = "pass",
  test = "LinDA",
  covariates = c("Age")
)

pval_mat <- res[["p.val"]][[g]][[subset_name]]
head(rownames(pval_mat))
eff_mat <- res[["eff.size"]][[g]][[subset_name]]
eff <- eff_mat[, rep_col]

sum(marker_idx_h5 %in% rownames(pval_mat))
setdiff(marker_idx_h5, rownames(pval_mat))

p <- pval_mat[, rep_col]
padj <- p.adjust(p, method = "fdr")
res_df <- data.frame(feature=names(eff), pval=p, padj=padj, effect_size=eff)

detected <- names(padj)[padj < alpha]

cat("Detected taxa:", length(detected), "\n")
cat("True markers:", length(marker_idx_h5), "\n")
cat("True positives:", sum(detected %in% marker_idx_h5), "\n")
cat("False positives:", sum(!detected %in% marker_idx_h5), "\n")
cat("Manual power:", mean(marker_idx_h5 %in% detected), "\n")

```

# Prepare the benchmark groups

SIMBA stores one HDF5 group per simulation setting. The chunks below identify all groups and restrict the analysis to the intended number of simulation repeats.

```{r prepare-groups}
temp <- rhdf5::h5ls(sim.implant, recursive = FALSE)
groups <- setdiff(temp$name, c("original_data", "simulation_parameters"))

# Keep abundance-scale groups with the single prevalence setting used here.
groups.sub <- groups[grepl("^ab[0-9]+_prev1_rep[0-9]+$", groups)]

# Restrict to the requested number of simulation repeats.
groups.sub <- groups.sub[
  as.integer(stringr::str_match(groups.sub, "^ab[0-9]+_prev1_rep([0-9]+)$")[, 2]) <= REPS_USED
]

rm(temp, groups)
```

# Benchmark multiple differential abundance methods

The benchmark runs one method at a time across all simulated groups and then aggregates performance metrics across repeats.

## Run the benchmark

The normalization strategy is chosen separately for each method. Calibration tables are also extracted where available, as they are needed later for detectability-boundary analyses.

```{r benchmark-many-methods, message=FALSE, warning=FALSE, results='hide'}

test_names <- c(
  "LinDA",
  "metagenomeSeq", 
  "ANCOMBC2", "ANCOMBC",
  "DESeq2", "lm", "edgeR"
)
  
force_DA_tests <- FALSE
filename = file.path(simh5.loc, paste0("BC_", param_set, "_eval_summary.rds"))
if (!file.exists(filename) || force_DA_tests) {
  
  # Modify this 
  n_workers <- n_cpus
  
  if (n_workers > 1) {
    future::plan(future::multisession, workers = n_workers)
  } else {
    future::plan("sequential")
  }
  
  method_results_BC <- test_names %>%
    set_names() %>%
    purrr::map(function(test_name) {
      # Method-specific normalization
      norm_method <- dplyr::case_when(
        test_name == "lm" ~ "clr_halfmin",
        test_name == "wilcoxon" ~ "TSS",
        test_name == "LinDA" ~ "TSS",
        TRUE ~ "pass"
      )
      
      # Covariate adjustment where relevant
      covariates <- NULL
      if (test_name %in% c("LinDA", "metagenomeSeq", "lm", "ANCOMBC", "ANCOMBC2", "DESeq2", "edgeR")) {
        covariates <- c("Age")
      }
      
      # Some methods do not currently return effect sizes in the patched workflow
      return_calib <- TRUE
      
      simbaR::run_method_eval_parallel(
        test_name = test_name,
        groups.sub = groups.sub,
        sim.implant = sim.implant,
        subset_size = subset_size,
        ab.scale = ab.scale,
        alpha = ALPHA,
        norm = norm_method,
        covariates = covariates,
        seed = TRUE,
        show_progress = TRUE,
        simba.loc = NULL,
        return_calib = return_calib
      )
    })
  saveRDS(method_results_BC, file = filename)
} else {
  method_results_BC <- readRDS(file = filename)
}

```


## Summarize benchmark performance

```{r benchmark-summary}
master_evals_BC <- dplyr::bind_rows(purrr::map(method_results_BC, "evals"))

power_combined_benchmark_BC <- master_evals_BC %>%
  group_by(test, ab_scale, prev_index) %>%
  summarise(
    mean_power = mean(power, na.rm = TRUE),
    sd_power = sd(power, na.rm = TRUE),
    mean_fdr = mean(empirical_fdr, na.rm = TRUE),
    sd_fdr = sd(empirical_fdr, na.rm = TRUE),
    mean_precision = mean(precision, na.rm = TRUE),
    mean_auroc = mean(auroc, na.rm = TRUE),
    mean_abs_eff_tp2 = mean(mean_abs_eff_tp, na.rm = TRUE),
    sd_abs_eff_tp = sd(mean_abs_eff_tp, na.rm = TRUE),
    mean_abs_eff_marker2 = mean(mean_abs_eff_marker, na.rm = TRUE),
    sd_abs_eff_marker = sd(mean_abs_eff_marker, na.rm = TRUE),
    n_reps = n(),
    .groups = "drop"
  )

datatable(power_combined_benchmark_BC)
```

## Visualize benchmark performance

These plots provide the benchmark overview used to select methods for the downstream detectability analyses.

```{r benchmark-viz, fig.width=8, fig.height=10}
plot_df <- power_combined_benchmark_BC %>%
  filter(test %in% test_names) %>%
  mutate(test = factor(test, levels = test_names))

p_fdr <- ggplot(plot_df, aes(x = ab_scale, y = mean_fdr, color = test, group = test)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(
      ymin = pmax(0, mean_fdr - sd_fdr),
      ymax = pmin(1, mean_fdr + sd_fdr)
    ),
    width = 0.08,
    alpha = 0.6
  ) +
  #geom_hline(yintercept = ALPHA, linetype = 2) +
  geom_hline(yintercept = 0.20, linetype = 2) +
  scale_x_continuous(limits = c(0.5, 10), breaks = seq(1, 10)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Implanted effect size",
    y = "FDR",
    color = "Method"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

p_power <- ggplot(plot_df, aes(x = ab_scale, y = mean_power, color = test, group = test)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(
      ymin = pmax(0, mean_power - sd_power),
      ymax = pmin(1, mean_power + sd_power)
    ),
    width = 0.08,
    alpha = 0.6
  ) +
  geom_hline(yintercept = 0.75, linetype = 2) +
  #geom_hline(yintercept = 0.4, linetype = 2) +
  scale_x_continuous(limits = c(0.5, 10), breaks = seq(1, 10)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Implanted effect size",
    y = "Power",
    color = "Method"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

BC_summary_plot <- ((p_power + labs(tag = "a")) / (p_fdr + labs(tag = "b"))) +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "bottom",
    plot.tag = element_text(size = 20, face = "bold")
  )

BC_summary_plot

ggsave(filename = file.path(out.loc, paste0("BC_", param_set, "_summary.pdf")), 
       plot = BC_summary_plot,
       device = cairo_pdf,
       width = 8,
       height = 10)
ggsave(filename = file.path(out.loc, paste0("BC_", param_set, "_power.pdf")), plot = p_power, width = 8, height = 5)
ggsave(filename = file.path(out.loc, paste0("BC_", param_set, "_fdr.pdf")), plot = p_fdr, width = 8, height = 5)
```

# Build the calibration table

The calibration analysis uses method-specific p-values and effect sizes extracted from the benchmarked simulations.

```{r calibration-analysis}
master_calib_BC <- dplyr::bind_rows(purrr::map(method_results_BC, "calib_df")) %>%
  mutate(
    detected = as.logical(padj < ALPHA),
    is_tp = as.logical(detected & is_marker),
    is_fp = as.logical(detected & !is_marker)
  )
```

```{r debug, eval=FALSE, include=FALSE}

# Read from HDF5

xgrp <- rhdf5::h5read(sim.implant, "ab1_prev1_rep1")
xfeat <- xgrp$features
rownames(xfeat) <- xgrp$feature_names
colnames(xfeat) <- xgrp$sample_names

xmeta <- rhdf5::h5read(sim.implant, "original_data/metadata")
xmeta <- as.data.frame(xmeta, stringsAsFactors = FALSE)
rownames(xmeta) <- xmeta[["__sample_id"]]
xmeta["__sample_id"] <- NULL
newlabel <- xmeta$DiseaseStatus.binary
newlabel[newlabel == 1] <- -1
newlabel[newlabel == 2] <- 1
xmeta$DiseaseStatus.binary <- newlabel

isTRUE(all.equal(colnames(xfeat), rownames(xmeta)))

xotu <- "ext_mOTU_v3_17336"

y <- feat[xotu, rownames(xmeta)]
predf <- cbind(meta, raw = as.numeric(y))

y[y==0] <- 0.5
y <- clr(y)

xdf <- cbind(xmeta, x = as.numeric(y))

xfit <- lm(x ~ DiseaseStatus.binary + Age, data = xdf)
broom::tidy(xfit)

# Test from phyloseq

tab_analysis <- as(otu_table(ps2), "matrix")
#tab_analysis <- tab_analysis[, rownames(meta)]
tab_analysis[tab_analysis == 0] <- 0.5
tab_analysis <- as.data.frame(apply(tab_analysis, 1, compositions::clr))
ydf = cbind(xmeta, y=as.numeric(tab_analysis[,"ext_mOTU_v3_17336"]))
yfit <- lm(y ~ DiseaseStatus.binary + Age, data = ydf)
broom::tidy(yfit)


zdf <- inner_join(inner_join(predf, xdf), ydf)

```


# Select methods for interpretability analysis

Only methods that jointly satisfy the chosen benchmark criteria are carried forward into the detectability-boundary plots. We have relaxed the criteria to be: above 40% power and below 15% FDR.

```{r select-methods}
good_ones <- power_combined_benchmark_BC %>%
  filter(mean_power >= 0.75 & mean_fdr <= 0.20) %>%
  group_by(test) %>%
  filter(rank(ab_scale, ties.method = "first") == 1) %>%
  arrange(ab_scale, test)
datatable(good_ones)

methods_keep <- sort(good_ones$test)
methods_keep
```

# Run the real-data differential abundance analyses

The next step is to run the corresponding differential abundance methods on the real breast cancer case/control data. These results are then compared with the simulation-derived detectability boundaries.

```{r real-data-DA-analysis, message=FALSE, warning=FALSE, results='hide'}

# prepare the tables
ps2 <- filter_taxa_by_prevalence(ps, MIN_DISCOVERY_PREVALENCE)
ps2rel <- transform_sample_counts(ps2, function(x) x / sum(x))
real_feat_relab <- as(otu_table(ps2rel), "matrix")
real_feat_relab <- as.data.frame(real_feat_relab[, rownames(meta)])

# Run DA test on real data

force_DA_tests <- FALSE

param_set2 <- paste("alpha", ALPHA, "prev", MIN_DISCOVERY_PREVALENCE, "reps", REPS_USED, sep = "_")
filename = file.path(simh5.loc, paste0("BC_", param_set2, "_real_data.rds"))

if (!file.exists(filename) || force_DA_tests) {
  
  # DESeq2
  
  sample_data(ps2)[["DiseaseStatus.binary"]] <- factor(sample_data(ps2)[["DiseaseStatus.binary"]],
                                                       levels=c(1,2),
                                                       ordered=FALSE)
  DES = phyloseq_to_deseq2(ps2, as.formula("~ Age + DiseaseStatus.binary"))
  dds <- estimateSizeFactors(DES, type = "poscounts")
  dds <- DESeq(dds,
               fitType = "parametric",
               test = "Wald",
               quiet = TRUE,
               minReplicatesForReplace = Inf
  )
  res_DESeq2 <- as.data.frame(results(dds, contrast = c("DiseaseStatus.binary", c(1, 2)), cooksCutoff = TRUE))
  
  # Encode as 1/2 as 0/1
  
  meta$DiseaseStatus.binary <- meta$DiseaseStatus.binary - 1
  sample_data(ps2) <- meta
  
  # ANCOMBC2
  res_ancombc2 <- ancombc2(
    data = ps2,
    fix_formula = "DiseaseStatus.binary + Age",
    rand_formula = NULL,
    p_adj_method = "fdr",
    prv_cut = MIN_DISCOVERY_PREVALENCE,
    lib_cut = 1000,
    alpha = ALPHA
  )
  # Filter for sensitivity analysis
  res_ancombc2$res <- res_ancombc2$res %>%
    dplyr::mutate(p_DiseaseStatus.binary = replace(p_DiseaseStatus.binary, passed_ss_DiseaseStatus.binary != TRUE, 1))
  
  # ANCOMBC
  res_ancombc <- ancombc(
    data = ps2,
    formula = "DiseaseStatus.binary + Age",
    p_adj_method = "fdr",
    pseudo = 1,
    prv_cut = MIN_DISCOVERY_PREVALENCE,
    lib_cut = 1000
  )
  
  # Re-encode 0/1 as -1/+1
  newlabel <- as.numeric(meta$DiseaseStatus.binary)
  newlabel[newlabel == 0] <- -1
  # newlabel[newlabel == 1] <- 1
  meta$DiseaseStatus.binary <- newlabel
  
  # lm on clr-transformed abundance with a half-minimum pseudocount
  tab_analysis <- as(otu_table(ps2), "matrix")
  pseudo <- min(tab_analysis[tab_analysis > 0]) / 2
  tab_analysis[tab_analysis == 0] <- pseudo
  tab_analysis <- as.data.frame(apply(tab_analysis, 1, compositions::clr))
  
  fit_one <- function(y, meta) {
    d <- cbind(meta, y = as.numeric(y))
    fit <- lm(y ~ DiseaseStatus.binary + Age, data = d)
    res <- broom.mixed::tidy(fit, effects = "fixed")
    res[res$term == "DiseaseStatus.binary", ]
  }
  
  res_lm <- apply(tab_analysis, MARGIN=2, FUN = fit_one, meta = meta) %>%
    bind_rows(.id = "OTU") %>%
    mutate(padj = p.adjust(p.value, method = "fdr"))
  
  # LinDA on relative abundance
  
  meta$DiseaseStatus.binary = factor(meta$DiseaseStatus.binary, levels=c(-1, 1))
  res_linda <- LinDA::linda(
    real_feat_relab,
    meta,
    formula = "~ DiseaseStatus.binary + Age",
    alpha = ALPHA,
    p.adj.method = "BH",
    type = "proportion"
  )
  
  # metagenomeSeq
  
  sample_data(ps2) <- meta # take the version with factors and -1/+1
  
  m <- phyloseq_to_metagenomeSeq(ps2)
  m2 <- filterData(m, present = nsamples(ps2) * MIN_DISCOVERY_PREVALENCE, depth = 1000)
  p <- cumNormStat(m2)
  m2 <- cumNorm(m2, p = p)
  pd <- pData(m2)
  
  mod1 <- model.matrix(~ DiseaseStatus.binary + Age, data = pd)
  res1 <- fitZig(m2, mod1, control = zigControl(maxit = 20, verbose = FALSE))
  res_metagenomeSeq <- MRcoefs(res1, number = nrow(res1), adjustMethod = "fdr")
  
  
  # Store raw p-values and effect sizes for downstream interpretability analyses
  p_lm_real <- as.numeric(res_lm$p.value)
  eff_lm_real <- as.numeric(res_lm$estimate)
  p_linda_real = res_linda$output$DiseaseStatus.binary$pvalue
  eff_linda_real = res_linda$output$DiseaseStatus.binary$log2FoldChange
  p_mgs_real <- res_metagenomeSeq$pvalues
  eff_mgs_real <- res_metagenomeSeq$DiseaseStatus.binary1
  p_ancombc_real <- res_ancombc$res$p_val$DiseaseStatus.binary
  eff_ancombc_real <- res_ancombc$res$lfc$DiseaseStatus.binary
  p_ancombc2_real <- res_ancombc2$res$p_DiseaseStatus.binary
  eff_ancombc2_real <- res_ancombc2$res$lfc_DiseaseStatus.binary
  
  p_deseq2_real = res_DESeq2$pvalue
  eff_deseq2_real = res_DESeq2$log2FoldChange
  
  
  names(p_lm_real) <- rownames(real_feat_relab)
  names(eff_lm_real) <- rownames(real_feat_relab)
  names(p_linda_real) <- rownames(real_feat_relab)
  names(eff_linda_real) <- rownames(real_feat_relab)
  names(p_mgs_real) <- rownames(real_feat_relab)
  names(eff_mgs_real) <- rownames(real_feat_relab)
  names(p_ancombc_real) <- rownames(real_feat_relab)
  names(eff_ancombc_real) <- rownames(real_feat_relab)
  names(p_ancombc2_real) <- res_ancombc2$res$taxon
  names(eff_ancombc2_real) <- res_ancombc2$res$taxon
  names(p_deseq2_real) <- rownames(real_feat_relab)
  names(eff_deseq2_real) <- rownames(real_feat_relab)
  
  real_results_BC <- list(
    DESeq2 = list(pval = p_deseq2_real, eff = eff_deseq2_real),
    lm = list(pval = p_lm_real, eff = eff_lm_real),
    LinDA = list(pval = p_linda_real, eff = eff_linda_real),
    ANCOMBC = list(pval = p_ancombc_real, eff = eff_ancombc_real),
    ANCOMBC2 = list(pval = p_ancombc2_real, eff = eff_ancombc2_real),
    metagenomeSeq = list(pval = p_mgs_real, eff = eff_mgs_real)
  )
  
  saveRDS(real_results_BC, file = filename)
} else {
  real_results_BC <- readRDS(file = filename)
}

```


```{r debug-linda, include=FALSE, eval=FALSE}
ps_rds <- "/projects/arumugam/people/xhf865/Camila/BreastCancer/phyloseq/ps.motus_raw.3.0.3.rds"
ps_orig <- readRDS(ps_rds)
ps <- filter_taxa_by_prevalence(ps_orig, MIN_DISCOVERY_PREVALENCE)
ps

meta <- sample_data(ps) %>%
  unclass() %>%
  as.data.frame() %>%
  select(participant_id,
         DNA_ID,
         DiseaseStatus.binary,
         Age
         )
meta$Individual_ID <- as.character(meta$participant_id)
meta$participant_id <- NULL
meta$DiseaseStatus.binary <- as.numeric(meta$DiseaseStatus.binary)
meta$Timepoint <- 1
rownames(meta) <- meta$DNA_ID
meta$DNA_ID <- NULL

sample_data(ps) <- meta

ps2 <- transform_sample_counts(ps, function(x) x / sum(x))
ps2
meta <- as(sample_data(ps2), "data.frame")
meta$DiseaseStatus.binary[meta$DiseaseStatus.binary == 0] = -1
meta$DiseaseStatus.binary[meta$DiseaseStatus.binary == 1] = 1
meta$DiseaseStatus.binary[meta$DiseaseStatus.binary == -1] = 0
meta$DiseaseStatus.binary = as.numeric(meta$DiseaseStatus.binary)
tab_analysis <- as(otu_table(ps2), "matrix")
tab_analysis <- tab_analysis[, rownames(meta)]
res_linda <- LinDA::linda(
  tab_analysis,
  meta,
  formula = "~ DiseaseStatus.binary + Age",
  alpha = ALPHA,
  p.adj.method = "BH",
  type = "proportion"
)
res_linda$output$DiseaseStatus.binary["meta_mOTU_v3_12389",]
```

# Build the abundance-stratified detectability analysis

```{r detectability-modeling}

force_boundary_modeling <- FALSE

param_set2 <- paste("alpha", ALPHA, "prev", MIN_DISCOVERY_PREVALENCE, "reps", REPS_USED, sep = "_")
filename = file.path(simh5.loc, paste0("BC_", param_set2, "_detectability_modeling.rds"))

if (!file.exists(filename) || force_boundary_modeling) {
  detectability_model_res <- compare_detectability_models_aic_cv(
    master_calib = master_calib_BC %>% filter(ab_scale > 1),
    methods = c("ANCOMBC", "ANCOMBC2", "LinDA", "lm"),
    target_col = "detected",
    good_col = "is_tp",
    model_type = "nonlinear"
  )
  saveRDS(detectability_model_res, file = filename)
} else {
  detectability_model_res <- readRDS(file = filename)
}

datatable(detectability_model_res$aic_long)
datatable(detectability_model_res$best_model_aic)
datatable(detectability_model_res$cv_summary_auc)
datatable(detectability_model_res$best_model_auroc)
datatable(detectability_model_res$best_model_auprc)

```


## Plot the real analysis results against abundance-stratified predictability boundaries

The plots below places taxa from real data against the simulation-derived 80% detection barrier.

```{r estimate-detection-thresholds-abundance}
real_prevalence <- rowMeans(feat > 0)

res_bound_abund_BC <- run_casecontrol_upper_bound_abundance(
  master_calib = master_calib_BC %>% dplyr::filter(ab_scale > 1),
  feat_real = real_feat_relab,
  methods = methods_keep,
  real_results_list = real_results_BC,
  analysis_subset_col = "is_marker",
  response_col = "detected",
  prevalence = real_prevalence,
  target_prob = 0.8,
  adjust = "fdr",
  alpha = 0.20,
  formula = good_bin ~
        s(effect, k = 4, bs = "ts") +
        s(prevalence, k = 4, bs = "ts") +
        s(abundance, k = 4, bs = "ts")
)
```

Let's visualize whether real data effect sizes are entering the detectable regions.

```{r plot-detection-thresholds, fig.width=8, fig.height=11}
p_bound_BC_model <- plot_effect_size_upper_bounds_abundance(
  res_bound = res_bound_abund_BC,
  methods = methods_keep
)
p_bound_BC_model
ggsave(
  filename = file.path(out.loc, paste0("BC_", param_set, "_eff_size_boundaries_real.pdf")),
  plot = p_bound_BC_model,
  device = cairo_pdf,
  width = 8,
  height = 9
)
```

Now let's visualize this together with the corresponding p-values, to put the ones crossing the boundary in context.

```{r plot-detection-thresholds-pval, fig.width=8, fig.height=11}
miami_width = 8
miami_height = 11
p_pseudomiami_BC <- plot_pseudomiami_detectability_abundance(
  res_bound = res_bound_abund_BC,
  methods = methods_keep,
  pval_ymax = 4,
  top_fraction = 0.5,
  gap_fraction = 0.05,
  figure_width = miami_width,
  figure_height = miami_height,
  failed_label = "Species failing the association test",
  positive_label = "Positive taxon",
  boundary_label = "80% detectability boundary",
  annotation_method = "ANCOMBC2",
  annotation_stratum = "High",
  annotation_x_lower = 0.8,
  annotation_x_upper = 0.4,
  annotation_angle = 20
)
p_pseudomiami_BC
ggsave(
  filename = file.path(out.loc, paste0("BC_", param_set, "_pseudo_miami.pdf")),
  plot = p_pseudomiami_BC,
  device = cairo_pdf,
  width = miami_width,
  height = miami_height
)
```


```{r knitting, eval=FALSE, include=FALSE}
rmarkdown::render(
  "BC_detectability_analysis.Rmd",
  output_file = "BC_detectability_analysis.publish.html",
  output_dir = "."
)
```

# Session information

Session information is printed below to document package versions and the computational environment used for the analysis.

```{r session-info}
sessionInfo()
```
